---
title: "Public health and economic problems caused by severe weather events"
author: "Jonas Wiorek"
date: "19 Mar 2015"
output: html_document
---
##Synopsis
In this report we describe the public health and economic problems caused by storms and other severe weather events across the United States from 1950 to November 2011. In particular, we describe the types of events which are most harmful with respect to population health, and the types of events which have the greatest economic consequences for communities and municipalities. It is a key concern to prevent such outcomes to the extent possible.
For this purpose the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database is studied. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

##Data Processing
The data for this study come from U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database in the form of a comma-separated-value file.
```{r, echo=TRUE, cache=TRUE}
if(!file.exists("data")){
        dir.create("data")
}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileUrl, destfile="./data/repdata-data-StormData.csv.bz2", method="curl")
stormData <- read.csv(bzfile("./data/repdata-data-StormData.csv.bz2"))
```


## Results
```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(xtable)
library(gridExtra)
```

### Most Harmful Population Health Events
The tables below show the top ten most harmful severe weather events with respect to population health in form of fatalities respectively injures. 
```{r, echo=TRUE, results='asis'}
# A data frame is created that summarize the number of fatatlities and injuries
# per event type
eventFatalitiesInjuries <- summarize(group_by(stormData, EVTYPE), 
                                     fatalities = 
                                             as.integer(sum(FATALITIES, na.rm=TRUE)),
                                     injuries = 
                                             as.integer(sum(INJURIES, na.rm =TRUE)))

# A data frame is created which order the events in number of fatalities in
# descending order. The top ten is exctracted.
fatalitiesOrdered <- arrange(eventFatalitiesInjuries, desc(fatalities))
fatalitiesOrderedFiltered <- fatalitiesOrdered[1:10,]
xt1 <- xtable(fatalitiesOrderedFiltered[,1:2])
print.xtable(xt1, type = "html")

# A data frame is created which order the events in number of injuries in
# descending order. The top ten is exctracted.
injuriesOrdered <- arrange(eventFatalitiesInjuries, desc(injuries))
injuriesOrderedFiltered <- injuriesOrdered[1:10,]
xt2 <- xtable(injuriesOrderedFiltered[,c(1,3)])
print.xtable(xt2, type = "html")

g1 <- ggplot(data=fatalitiesOrderedFiltered,aes(x=EVTYPE,y=fatalities)) +
  geom_bar( stat="identity") +
  scale_x_discrete(limits = fatalitiesOrderedFiltered$EVTYPE) +
  theme(text = element_text(size=12), axis.text.x = element_text(angle=90))

g2 <- ggplot(data=injuriesOrderedFiltered,aes(x=EVTYPE,y=injuries)) +
  geom_bar( stat="identity") +
  scale_x_discrete(limits = injuriesOrderedFiltered$EVTYPE) +
  theme(text = element_text(size=12), axis.text.x = element_text(angle=90))

#par(mfcol = c(1,2))
#grid.arrange(g1,g2)
```
The most harmful events in form of fatilities are `r fatalitiesOrderedFiltered$EVTYPE[1]` followed by `r fatalitiesOrderedFiltered$EVTYPE[2]` and `r fatalitiesOrderedFiltered$EVTYPE[3]`.

The most harmful events in form of injures are `r injuriesOrderedFiltered$EVTYPE[1]` followed by `r injuriesOrderedFiltered$EVTYPE[2]` and `r injuriesOrderedFiltered$EVTYPE[3]`.

### Events of Greatest Economic Consequences
The figures below show the top ten types severe weather events which have the greatest economic consequences with respect to property respectively crop damages.
```{r, echo=TRUE, results='asis'}
# Function to translate PROPDMGEXP and CROPDMGEXP to factors to multiple the values
# in columns PROPDMG and CROPDMG.
# The values in PROPDMGEXP and CROPDMGEXP could take several values. They
# are tranlated into a multiple factor according to the table
# "" -> 1 : The value in PROPDMG/CROPDMG is the actual value
# "+" -> 1 : The value in PROPDMG/CROPDMG is likely low but still the best estimate
# "-" -> 1 : The value in PROPDMG/CROPDMG is likely high but still the best estimate
# "?" -> 1 : The value in PROPDMG/CROPDMG is uncertain but still the best estimate
# "0-8" -> 10^0-8 : The value in PROPDMG/CROPDMG is multiplied with the value in PROPDMGEXP/CROPDMGEXP as the exponent of 10
# "[Hh]" -> 10^2 : The value in PROPDMG/CROPDMG is multiplied with 10^2
# "[Kk]" -> 10^3 : The value in PROPDMG/CROPDMG is multiplied with 10^3
# "[Mm]" -> 10^6 : The value in PROPDMG/CROPDMG is multiplied with 10^6
# "[B]" -> 10^9 : The value in PROPDMG/CROPDMG is multiplied with 10^9
exp2Factor <- function(.) {
        ifelse(. %in% "B",times <- 10^9,
        ifelse(. %in% c("M","m"), times <- 10^6, 
        ifelse(. %in% c("K","k"), times <- 10^3,
        ifelse(. %in% c("H","h"), times <- 10^2,
        ifelse(. %in% "1", times <- 10^1,
        ifelse(. %in% "2", times <- 10^2,
        ifelse(. %in% "3", times <- 10^3,
        ifelse(. %in% "4", times <- 10^4,
        ifelse(. %in% "5", times <- 10^5,
        ifelse(. %in% "6", times <- 10^6,
        ifelse(. %in% "7", times <- 10^7,
        ifelse(. %in% "8", times <- 10^8,
               times <- 1))))))))))))
}

# Two columns,PROPDMGFACTOR and CROPDMGFACTOR, are added to the storm data. The 
# two columns represents the multiplication factor of the property and crop 
# damage values respectivaley given in columns PROPDMG and CROPDMG for each event.
stormData <- mutate(stormData, PROPDMGFACTOR=exp2Factor(PROPDMGEXP), CROPDMGFACTOR=exp2Factor(CROPDMGEXP))

# A data frame is created that summarize the economical consequences of 
# property and crop damages per event type.
eventPropertyCrop <- summarize(group_by(stormData, EVTYPE), 
                                     property = sum(PROPDMG*PROPDMGFACTOR, na.rm=TRUE),
                                     crop = sum(CROPDMG*CROPDMGFACTOR, na.rm =TRUE))

# A data frame is created which order the events in descending order of economical
# consequences of damages of property. The top ten is exctracted.
propertyOrdered <- arrange(eventPropertyCrop, desc(property))
propertyOrderedFiltered <- propertyOrdered[1:10,]

# A data frame is created which order the events in descending order of economical
# consequences of damages of crop. The top ten is exctracted.
cropOrdered <- arrange(eventPropertyCrop, desc(crop))
cropOrderedFiltered <- cropOrdered[1:10,]

ggplot(data=propertyOrderedFiltered,aes(x=EVTYPE,y=property)) +
        geom_bar( stat="identity") +
        scale_x_discrete(limits = propertyOrderedFiltered$EVTYPE) +
        theme(text = element_text(size=12), axis.text.x = element_text(angle=90)) +
        xlab("Event Type") +
        ylab("Property Damage [$]")

ggplot(data=cropOrderedFiltered,aes(x=EVTYPE,y=crop)) +
        geom_bar( stat="identity") +
        scale_x_discrete(limits = cropOrderedFiltered$EVTYPE) +
        theme(text = element_text(size=12), axis.text.x = element_text(angle=90)) +
        xlab("Event Type") +
        ylab("Crop Damage [$]")
```

The most harmful events in form of economic consequences with respect to property damages are `r propertyOrderedFiltered$EVTYPE[1]` followed by `r propertyOrderedFiltered$EVTYPE[2]` and `r propertyOrderedFiltered$EVTYPE[3]`.

The most harmful events in form of economic consequences with respect to crop damages are `r cropOrderedFiltered$EVTYPE[1]` followed by `r cropOrderedFiltered$EVTYPE[2]` and `r cropOrderedFiltered$EVTYPE[3]`.

